A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2024-04-11 tidyindex.tex
Minor change in section 1

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Science, University of Texas at Austin, Austin,
Texas, USA\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{3} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{4} Guangdong Provincial Key
Laboratory of Interdisciplinary Research and Application for Data
Science, BNU-HKBU United International College, Zhuhai,
Guangdong, China\\ \textsuperscript{5} School of Mathematics and
Statistics, University of Melbourne, Melbourne, Victoria, Australia
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and the variability therein. Here we
discuss a modular data pipeline recommendation to assemble indexes. It
is universally applicable to index computation and allows investigation
of index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
usage of the pipeline framework: comparison of two different indexes
designed to monitor the spatio-temporal distribution of drought in
Queensland, Australia; the effect of dimension reduction choices on the
Global Gender Gap Index (GGGI) on countries' ranking; and how to
calculate bootstrap confidence intervals for the Standardized
Precipitation Index (SPI). The methods are supported by a new R package,
called \texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, sharp corners]}{\end{tcolorbox}}\fi

@@

To construct an index, experts typically start by defining a concept of
interest that requires measurement. This concept often lacks a direct
measurable attribute or can only be measured as a composite of various
processes, yet it holds social and public significance. To create an
index, once the underlying processes involved are identified, relevant
and available variables are then defined, collected, and combined using
statistical methods into an index that aims to measure the process of
interest. The construction process is often not straightforward, and
decisions need to be made, such as the selection of variables to be
included, which might depend on data availability and the statistical
definition of the index to be used, among others. For instance, the
indexes constructed from a linear combination of variables require to
decide the weight assigned to each variable. Some indexes have a spatial
and/or temporal component, and variables can be aggregated to different
spatial resolutions and temporal scales, leading to various indexes for
different monitoring purposes. Hence, all these decisions can result in
different index values and have different practical implications.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Science, University of Texas at Austin, Austin,
Texas, USA\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{3} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{4} Guangdong Provincial Key
Laboratory of Interdisciplinary Research and Application for Data
Science, BNU-HKBU United International College, Zhuhai,
Guangdong, China\\ \textsuperscript{5} School of Mathematics and
Statistics, University of Melbourne, Melbourne, Victoria, Australia
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and the variability therein. Here we
discuss a modular data pipeline recommendation to assemble indexes. It
is universally applicable to index computation and allows investigation
of index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
usage of the pipeline framework: comparison of two different indexes
designed to monitor the spatio-temporal distribution of drought in
Queensland, Australia; the effect of dimension reduction choices on the
Global Gender Gap Index (GGGI) on countries' ranking; and how to
calculate bootstrap confidence intervals for the Standardized
Precipitation Index (SPI). The methods are supported by a new R package,
called \texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, frame hidden]}{\end{tcolorbox}}\fi

@@

To construct an index, experts typically start by defining a concept of
interest that requires measurement. This concept often lacks a direct
measurable attribute or can only be measured as a composite of various
processes, yet it holds social and public significance. To create an
index, once the underlying processes involved are identified, relevant
and available variables are then defined, collected, and combined using
statistical methods into an index that aims to measure the process of
interest. The construction process is often not straightforward, and
decisions need to be made, such as the selection of variables to be
included, which might depend on data availability and the statistical
definition of the index to be used, among others. For instance, the
indexes constructed from a linear combination of variables require a
decision on the weight assigned to each variable. Some indexes have a
spatial and/or temporal component, and variables can be aggregated to
different spatial resolutions and temporal scales, leading to various
indexes for different monitoring purposes. Hence, all these decisions
can result in different index values and have different practical
implications.
